Beyond the Arc: How 3 point shooting has impacted winning in the modern NBA

### 2024-04-17 paper/paper/paper.qmd
Add files via upload

The three point line was introduced in the NBA in the 1979-80 season and has brought unprecedented balance to the game. Instead of the game being centered around the big man or athletic wings, the 3 point shot gave everybody on the team a new powerful weapon on offenses. Today, there are 230 players who averaged at least one made three pointer in the 2022-23 regular season. The estimand of this paper is to see if these 3 point shots help the team win games in the regular season.

This paper aims to estimate the association between three-point shooting measures and win probability using advanced statistical techniques after meticulously analyzing season-wise data. The findings highlighted the importance of three-point shooting efficiency as a predictor of team performance in modern basketball and how important it is to be a good three point shooting team for success in today's game.

Understanding this link has far-reaching ramifications for numerous parties, including coaches, players, analysts, and team administration. This study adds to the ongoing debate in the NBA about strategic decision-making and player development by explaining the significance of three-point shooting in deciding game outcomes. Essentially, the question "Does three point shooting actually lead to team success?", is being asked.

I begin by exploring the underlying data, which will provide insights into the variables and metrics used in our analyses. Following that, I explain the complexities of the modelling approach, describing the reasoning behind the statistical techniques and justifying the methodology. Later, the results of the analysis are discussed, diving into the empirical findings and their implications for understanding the relationship between three-point shooting efficiency and NBA team success. Then  I have included a thorough conversation, contextualizing the findings within the larger landscape of basketball analytics and emphasizing significant takeaways from this research. Finally, I summarize the findings and propose paths for future research to further the understanding of this dynamic interaction.

# Data

In this study, I utilized data sourced from the NBA's official database (@cite-NBA) to examine the relationship between various team performance metrics and their corresponding winning percentages throughout the regular season. My analysis aims to uncover insights into the factors contributing to team success in the NBA, specifically 3 point shooting. The NBA's database serves as a complete repository of statistical information on teams, players, and games, providing valuable insights into the dynamics of professional basketball. Below, I provide a detailed overview of the data collection process, cleaning procedures, and subsequent analysis.

@@

All raw data utilized in this study was collected from the official NBA database (@cite-NBA), which aggregates all statistics on team performance, player performances, and game outcomes throughout the league's history. Since this study only focuses on the modern NBA, which is defined as the era from the 2014-15 season onward, the season wise data had to be individually downloaded and combined. Additionally, since the 2019-20 season was shortened due to the COVID-19 outbreak and an equal number of games was not played by all teams, we will not be considering that season for the sake of this study. This data is then stored in our raw_data2.xlsx file using @cite-writexl. This dataset comprises a wide array of variables, including team statistics such as points scored, rebounds, assists, turnovers, and various other performance metrics recorded over multiple seasons.

@@

Upon acquiring the raw data from nba.com (@cite-NBA), extensive cleaning procedures were implemented to ensure its suitability for analysis. This involved addressing missing or erroneous values, standardizing variable names, and aggregating data at the team level to facilitate meaningful comparisons across different seasons and teams. Additionally, I computed additional variables such as win percentages, and filtered the data to only include winning stats and 3 point stats using the *group_by* (@cite-groupby) package. This dataset is then stored in the analysis data as a csv file (@cite-writecsv).

@@

# Model

@@

The chosen Bayesian analysis model (@cite-rstanarm) aims to quantitatively assess the impact of 3-point field goals made on team wins and to provide insights into the significance of this performance metric in determining overall team success. By modeling the relationship between 3-point field goals made and wins using a normal distribution for the response variable, we acknowledge the continuous nature of team wins and seek to capture the variability in win counts across different levels of 3-point shooting proficiency. This approach allows for a nuanced understanding of how variations in 3-point performance contribute to variations in team outcomes, thereby elucidating the importance of 3-point shooting in the context of NBA team success.

# Results

@@

# Discussion

@@

## Future Research

@@

>>>>

The three point line was introduced in the NBA in the 1979-80 season (@cite-threepoint) and has brought unprecedented balance to the game. Instead of the game being centered around the big man or athletic wings, the 3 point shot gave everybody on the team a new powerful weapon on offenses. Today, there are 155 players who averaged at least one made three pointer in the 2022-23 regular season (@cite-threepoint2). The estimand of this paper is to see if these 3 point shots help the team win games in the regular season.

This paper explores the association between three-point shooting measures and win probability using advanced statistical techniques after meticulously analyzing season-wise data. The findings highlight the importance of three-point shooting efficiency as a predictor of team performance in modern basketball and how important it is to be a good three point shooting team for success in today's game.

Understanding this link has far-reaching ramifications for numerous parties, including coaches, players, analysts, and team administration. This study adds to the ongoing debate in the NBA about strategic decision-making and player development by explaining the significance of three-point shooting in deciding game outcomes. Essentially, the question “Does three point shooting actually lead to team success?”, is being asked.

This paper begins by exploring the underlying data in @sec-data, which provides insights into the variables and metrics used in the analyses. Following that, the complexities of the modelling approach are explained in @sec-model, including describing the reasoning behind the statistical techniques and justifying the methodology. The results of the analysis are discussed in @sec-results, diving into the empirical findings and their implications for understanding the relationship between three-point shooting eﬀiciency and NBA team success. @sec-discuss contextualizes the findings of this paper within the larger landscape of basketball analytics,emphasizing significant takeaways. Finally, the findings and proposed paths for future research are summarized in @sec-discuss2.

# Data {#sec-data}

This study, utilizes  data sourced from the NBA’s oﬀicial database (@cite-NBA) to examine the relationship between various team performance metrics and their corresponding winning percentages throughout the regular season. The analysis aims to uncover insights into the factors contributing to team success in the NBA, specifically 3 point shooting. The NBA’s database serves as a complete repository of statistical information on teams, players, and games, providing valuable insights into the dynamics of professional basketball. The below sections provide  a detailed overview of the data collection process, cleaning procedures, and subsequent analysis.

@@

All raw data utilized in this study was collected from the oﬀicial NBA database  (@cite-NBA), which aggregates statistics on team performance, player performances, and game outcomes throughout the league’s history. Since this study only focuses on the modern NBA, which is defined as the era from the 2014-15 season onward (@cite-modernera), the season wise data had to be individually downloaded and combined. Additionally, since the 2019-20 season was shortened due to the COVID-19 outbreak and an equal number of games was not played by all teams (@cite-shortseason), data from that season is excluded from this study to ensure consistency. The aggregated  dataset is then stored as the raw_data2.xlsx file using the *writexl* package (@cite-writexl). This dataset comprises data from 30 teams over 8 seasons, including team statistics such as points scored, rebounds, assists, turnovers, and various other performance metrics recorded over multiple seasons.

@@

Upon acquiring the raw data from NBA.com(@cite-NBA), extensive cleaning procedures were implemented to ensure its suitability for analysis. This involved addressing missing or erroneous values, standardizing variable names, and aggregating data at the team level to facilitate meaningful comparisons across different seasons and teams. Additionally, I computed additional variables such as win percentages, and filtered the data to only include winning stats and 3 point stats using the *group_by* (@cite-groupby) package. This dataset is then stored in the analysis data as a csv file (@cite-writecsv).

@@

# Model {#sec-model}

@@

The chosen Bayesian analysis model (@cite-rstanarm) aims to quantitatively assess the impact of 3-point field goals made on team wins and to provide insights into the significance of this performance metric in determining overall team success. By modeling the relationship between 3-point field goals made and wins using a normal distribution for the response variable, we acknowledge the continuous nature of team wins and seek to capture the variability in win counts across different levels of 3-point shooting proficiency. This approach allows for a nuanced understanding of how variations in 3-point performance contribute to variations in team outcomes, thereby elucidating the importance of 3-point shooting in the context of NBA team success. Simply put, teams that excel in scoring 3-point shots are more likely to outscore their opponents and win the games.

# Results {#sec-results}

@@

# Discussion {#sec-discuss}

@@

## Future Research {#sec-discuss2}

@@

{{< pagebreak >}}
